GEE models

### 2016-06-17 doc/manuscript.Rmd
Removed Zinman (as he requested) from author list
Made edits as per Ravi's edits

<!-- First, middle initial, last, highest academic degree -->
Author: Luke W. Johnston (1), MSc; Stewart B. Harris (2), MD; Ravi Retnakaran
(3,4), MD; Bernard Zinman (3,4), MD; Adria Giacca (5), PhD; Zhen Liu (1), PhD;
Richard P. Bazinet (1), PhD; and Anthony J. Hanley (1,6), PhD

@@

At each examination, an 8-12 hour fasting blood sample was draw from each
participant, followed by a 75g oral glucose tolerance test (OGTT) with a 30
minute and 2 hour blood draw. All blood samples were processed and frozen at
-70°C. Alanine aminotransferase (ALT) was measured using standard laboratory
procedures. Cholesterol, HDL, and triacylglycerides (TAG) were measured using Roche
Modular's enzymatic colorimetric tests (Mississauga, ON).  Both specific insulin
and glucose were derived from the OGTT at fasting, 30 minute, and 2 hour time
points. Specific insulin was measured using the Elecsys 1010 (Roche Diagnostics,
Basel, Switzerland) immunoassay analyzer and electrochemiluminescence
immunoassay. This assay shows 0.05% cross-reactivity to intact human pro-insulin
and the Des 31,32 circulating split form (Linco Res. Inc), and has a CV of 9.3%.
Glucose was determined using an enzymatic hexokinase (Roche Modular, Roche
Diagnostics) with a detection range of 0.11 (2 mg/dL) to 41.6 mmol/L.  The
inter-assay %CV is <1.1% and intra-assay %CV is < 1.9%.  All assays were
performed at the Banting and Best Diabetes Centre Core Lab at Mt Sinai Hospital.
Impaired fasting glucose (IFG), impaired glucose tolerance (IGT), and diabetes
were categorized using the 2006 WHO criteria [@WHO2006].

@@

For the primary analysis, generalized estimating equation (GEE) models 
[@Zeger1986a] were used to determine the longitudinal associations between the 
outcome variables and the predictor variables. Given the longitudinal design 
an auto-regressive of order 1 (AR1) correlation matrix was chosen for the GEE models, 
though other matrices (eg. exchangeable) had similar fit (data not shown). GEE
is well suited to longitudinal cohort studies given it's capacity to handle
missed visits. The predictor variables and continuous covariates were scaled
(mean centered and standardized). The NEFA variables were classified as
*time-independent* (held constant) as they were measured only at the baseline
visit, while the outcome variables and covariates were set as *time-dependent*.
No imputation was conducted on missing values.

@@

Three latent classes were extracted from LCMM, with `r lcmm.groups$Freq[2]` 
participants in the high, `r lcmm.groups$Freq[1]` in the middle, and 
`r lcmm.groups$Freq[3]`
in the low beta-cell function groups. The trajectories of the these groups are
shown in `r cite_sf('lcmm')`. These groups were used as the beta-cell function
response variables in the PLS-DA analysis. The PLS-DA results show a very poor
discriminatory ability of the NEFA composition in classifying participants into
the correct beta-cell function latent class. Only 
`r misclass.n.percent.bcf$NPct[2]` 
participants were correctly classified, primarily into the
middle group. The specific NEFA composition was not able to accurately classify
participants who had low or high beta-cell function, however, those individuals 
classified into the low beta-cell function group tended to have patterns of NEFA 
composition higher in 14:0, 16:0, 14:1n-7, and 16:1n-7.
`r cite_sf('plsda_loadings')` shows the loadings of individual NEFA by the
extracted component and `r cite_sf('plsda_groups')` shows the clustering of
the components by ISSI-2 latent class. The PLS-DA analysis using conversion to
or maintenance of dysglycemia status over the 6 years as the response variable showed similar
poor discriminatory ability, though better than the beta-cell function
trajectory results. While `r misclass.n.percent.dysgly$NPct[2]` participants 
were classified correctly based on dysglycemia status, this was almost entirely
due to classifying correctly into the non-dysglycemia status. Examining the
clustering of the NEFA composition showed poor discrimination between groups
(see `r cite_sf('plsda_groups_dysgly')` and 
`r cite_sf('plsda_loadings_dysgly')`).

@@

Biologically, in free-living populations chronic elevation of NEFA may be 
mediating it's association with metabolic outcomes through TAG. In normal
metabolism, NEFA enters the
liver and assists in the production of TAG that is to be processed into very-low
density lipoproteins (VLDL) [@Nielsen2012a]. As such, higher NEFA may contribute 
to hypertriglyceridemia, which is a known risk factor for diabetes. In the
sensitivity analysis adjusting for TAG, all associations with beta-cell function
were attenuated, suggesting that NEFA may in fact be mediating it's association
with beta-cell dysfunction through higher TAG.

There is substantial experimental evidence highlighting the role of increased 
NEFA and the subsequent increase in insulin resistance via impairment of insulin 
signaling cascades, as reviewed in previously published articles 
[@Martins2012a;@Ebbert2013a;@Capurso2012a]. However, in this
longitudinal analysis, we saw no association of any individual or total
NEFA with hepatic (1/HOMA-IR) or whole-body insulin sensitivity (ISI). 
There are several possible explanations for these findings. Previous studies
done *in vivo* were mostly short term infusion protocols
[@Xiao2009a;@Xiao2006;@Szendroedi2012a;@Kehlenbrink2012a] or pharmocologic
[@Daniele2014a;@Liang2013a] trials and previous observational research was generally
cross-sectional or correlational
[@Miller2012a;@Johns2014a;@Bush2012a;@Ilyasova2010a]. The short time periods of
these studies and the experimentally-induced
elevations in NEFA concentrations make it difficult to assess the causal role
in free-living environments over long time frames. Moreover, some studies
have found null or weak associations of NEFA and IR 
[@Gray2013a;@Magkos2012a;@Kehlenbrink2012a;@Johns2014a], suggesting that either
NEFA has only a minor role in IR or that NEFA is the effect, but not cause, of
the underlying IR. Adipose tissue, which release NEFA during fasting, can become
insulin resistant and thus lipolysis of intracellular TAG by hormone-sensitive 
lipase may not respond to the inhibitory action of insulin. As such, IR may 
develop before elevations in blood NEFA. There is also evidence to suggest that 
insulin may be involved in NEFA uptake [@Ramos-Roman2012a], further supporting 
the hypothesis that elevated NEFA may be a consequence of higher IR. However, we 
could not examine this potential hypothesis as NEFA was only collected at the
baseline visit.

>>>>

<!-- First, middle initial, last, highest academic degree -->
Author: Luke W. Johnston (1), MSc; Stewart B. Harris (2), MD; Ravi Retnakaran
(3,4), MD; Adria Giacca (5), PhD; Zhen Liu (1), PhD;
Richard P. Bazinet (1), PhD; and Anthony J. Hanley (1,6), PhD

@@

At each examination, an 8-12 hour fasting blood sample was drawn from each
participant, followed by a 75g oral glucose tolerance test (OGTT) with a 30
minute and 2 hour blood draw. All blood samples were processed and frozen at
-70°C. Alanine aminotransferase (ALT) was measured using standard laboratory
procedures. Cholesterol, HDL, and triacylglycerides (TAG) were measured using Roche
Modular's enzymatic colorimetric tests (Mississauga, ON).  Both specific insulin
and glucose were derived from the OGTT at fasting, 30 minute, and 2 hour time
points. Specific insulin was measured using the Elecsys 1010 (Roche Diagnostics,
Basel, Switzerland) immunoassay analyzer and electrochemiluminescence
immunoassay. This assay shows 0.05% cross-reactivity to intact human pro-insulin
and the Des 31,32 circulating split form (Linco Res. Inc), and has a CV of 9.3%.
Glucose was determined using an enzymatic hexokinase (Roche Modular, Roche
Diagnostics) with a detection range of 0.11 (2 mg/dL) to 41.6 mmol/L.  The
inter-assay %CV is <1.1% and intra-assay %CV is < 1.9%.  All assays were
performed at the Banting and Best Diabetes Centre Core Lab at Mt Sinai Hospital.
Impaired fasting glucose (IFG), impaired glucose tolerance (IGT), and diabetes
were categorized using the 2006 WHO criteria [@WHO2006].

@@

For the primary analysis, generalized estimating equation (GEE) models 
[@Zeger1986a] were used to determine the longitudinal associations between the 
outcome variables and the predictor variables. Given the longitudinal design 
an auto-regressive of order 1 (AR1) correlation matrix was chosen for the GEE models, 
though other matrices (eg. exchangeable) had similar fit (data not shown). GEE
is well suited to longitudinal cohort studies given its capacity to handle
missed visits. The predictor variables and continuous covariates were scaled
(mean centered and standardized). The NEFA variables were classified as
*time-independent* (held constant) as they were measured only at the baseline
visit, while the outcome variables and covariates were set as *time-dependent*.
No imputation was conducted on missing values.

@@

Three latent classes were extracted from LCMM, with `r lcmm.groups$Freq[2]` 
participants in the high, `r lcmm.groups$Freq[1]` in the middle, and 
`r lcmm.groups$Freq[3]`
in the low beta-cell function groups. The trajectories of these groups are
shown in `r cite_sf('lcmm')`. These groups were used as the beta-cell function
response variables in the PLS-DA analysis. The PLS-DA results show a very poor
discriminatory ability of the NEFA composition in classifying participants into
the correct beta-cell function latent class. Only 
`r misclass.n.percent.bcf$NPct[2]` 
participants were correctly classified, primarily into the
middle group. The specific NEFA composition was not able to accurately classify
participants who had low or high beta-cell function, however, those individuals 
classified into the low beta-cell function group tended to have patterns of NEFA 
composition higher in 14:0, 16:0, 14:1n-7, and 16:1n-7.
`r cite_sf('plsda_loadings')` shows the loadings of individual NEFA by the
extracted component and `r cite_sf('plsda_groups')` shows the clustering of
the components by ISSI-2 latent class. The PLS-DA analysis using conversion to
or maintenance of dysglycemia status over the 6 years as the response variable showed similar
poor discriminatory ability, though better than the beta-cell function
trajectory results. While `r misclass.n.percent.dysgly$NPct[2]` participants 
were classified correctly based on dysglycemia status, this was almost entirely
due to classifying correctly into the non-dysglycemia status. Examining the
clustering of the NEFA composition showed poor discrimination between groups
(see `r cite_sf('plsda_groups_dysgly')` and 
`r cite_sf('plsda_loadings_dysgly')`).

@@

Biologically, in free-living populations chronic elevation of NEFA may be 
mediating its association with metabolic outcomes through TAG. In normal
metabolism, NEFA enters the
liver and assists in the production of TAG that is to be processed into very-low
density lipoproteins (VLDL) [@Nielsen2012a]. As such, higher NEFA may contribute 
to hypertriglyceridemia, which is a known risk factor for diabetes. In the
sensitivity analysis adjusting for TAG, all associations with beta-cell function
were attenuated, suggesting that NEFA may in fact be mediating its association
with beta-cell dysfunction through higher TAG.

There is substantial experimental evidence highlighting the role of increased 
NEFA and the subsequent increase in insulin resistance via impairment of insulin 
signaling cascades, as reviewed in previously published articles 
[@Martins2012a;@Ebbert2013a;@Capurso2012a]. However, in this
longitudinal analysis, we saw no association of any individual or total
NEFA with hepatic (1/HOMA-IR) or whole-body insulin sensitivity (ISI). 
There are several possible explanations for these findings. Previous studies
done *in vivo* were mostly short term infusion protocols
[@Xiao2009a;@Xiao2006;@Szendroedi2012a;@Kehlenbrink2012a] or pharmacologic
[@Daniele2014a;@Liang2013a] trials and previous observational research was generally
cross-sectional or correlational
[@Miller2012a;@Johns2014a;@Bush2012a;@Ilyasova2010a]. The short time periods of
these studies and the experimentally-induced
elevations in NEFA concentrations make it difficult to assess the causal role
in free-living environments over long time frames. Moreover, some studies
have found null or weak associations of NEFA and IR 
[@Gray2013a;@Magkos2012a;@Kehlenbrink2012a;@Johns2014a], suggesting that either
NEFA has only a minor role in IR or that NEFA is the effect, but not cause, of
the underlying IR. Adipose tissue, which release NEFA during fasting, can become
insulin resistant and thus lipolysis of intracellular TAG by hormone-sensitive 
lipase may not respond to the inhibitory action of insulin. As such, IR may 
develop before elevations in blood NEFA. There is also evidence to suggest that 
insulin may be involved in NEFA uptake [@Ramos-Roman2012a], further supporting 
the hypothesis that elevated NEFA may be a consequence of higher IR. However, we 
could not examine this potential hypothesis as NEFA was only collected at the
baseline visit.
